A note on the instability and degeneracy of deep learning models

### 2016-11-30 writing/note.Rmd
edits from dr. nordman

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models). Section \ref{instability-results} establishes formal results regarding the propriety of FSFS models with regard to stability. A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$. The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential 
family models by @schweinberger2011instability. For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes. As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability. The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models. The main results here concern a general measure of model instability, appropriate accross the whole FSFS model class. This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$. In this case, a one component change in the data sequence may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.

@@

Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsmbol \theta}_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. (Note for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$; in this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

Consider two models with "deep architechture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast]. Let $M$ denote the number of hidden layers included in the model and $N_{(H,1)}, \dots, N_{((H,M))}$ the number of hidden variables within each hidden layer. Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ be random variables corresponding to the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilitic model. Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.

@@

This definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g. RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \ref{discrete-exponential-family-models}) considered there.

S-unstable FSFS model seqences are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.

>>>>

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models). Section \ref{instability-results} establishes formal results regarding the propriety of FSFS models with regard to stability. A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$. The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential 
family models by @schweinberger2011instability. For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes. As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability. The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models. The main results here concern a general measure of model instability, appropriate across the whole FSFS model class. This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$. In this case, a one component change in the data sequence may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.

@@

Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. (Note for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$; in this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast]. Let $M$ denote the number of hidden layers included in the model and $N_{(H,1)}, \dots, N_{((H,M))}$ the number of hidden variables within each hidden layer. Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ be random variables corresponding to the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model. Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.

@@

This definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing non-exponential family models (e.g. RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \ref{discrete-exponential-family-models}) considered there.

S-unstable FSFS model sequences are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.
